### 2020-08-03 paper/sex_selection_spacing-ver4.tex
Edit Abstract and Introduction

\date{July 2020}

@@

% Demography abstract
\noindent 
Using four rounds of India's National Family and Health Surveys and a
competing-risk hazard model, I show that Hindu women's average birth intervals
increased over the last four decades for all education groups. The most
significant increases are among the women most likely to use sex
selection. Despite the rise in average intervals, the likelihood of very
short spacing did not change substantially. Hence, the increases come
predominantly from the longer birth intervals getting even longer. As a
result of the longer spacing, fertility rates significantly
overestimated how fast cohort fertility fell. Although cohort fertility
and the fertility rate have started to converge, the cohort fertility is
still substantially higher than the fertility rate. Furthermore, cohort
fertility is still at or above replacement level for all but the
best-educated urban women. Finally, infant mortality risk has declined
substantially over time for all groups, but fastest for the lower
education groups, who are now close to the level of women with the most
education. Short birth spacing is still associated with higher
mortality, although the difference is small for the best-educated women.
There is no evidence that the increasing use of sex selection is
associated with higher infant mortality risk.

\noindent JEL: J1, O12, I1
\noindent Keywords: India, prenatal sex determination, censoring, competing risk, non-proportional hazard
\end{abstract}

@@

% Questions addressed and why valuable
The main question here is: how did birth spacing respond to these changes, especially 
the spread of sex selection? 
The motivation is twofold.
First, prior research has failed to appreciate that each abortion increases the
interval between births by six to twelve months.%
\footnote{
The increase consists of three parts. 
First, after an abortion, the uterus needs at least two menstrual cycles to recover, 
or the likelihood of spontaneous abortion increases substantially \citep{zhou00b}. 
Second, the waiting time to conception is one to six months. 
Finally, sex determination tests are reliable only from three months of gestation. 
}
Hence, the growing use of sex-selective abortions may substantially increase birth spacing, 
but we know little about how much. 
Second, more female education, higher household income, and low and declining female labor 
force participation all likely impact birth spacing. 
The combined changes in birth spacing may, therefore, outpace what we have observed in 
other countries.

Studying birth spacing contributes to our understanding of fertility decisions, but, equally 
important, birth spacing affects the reliability of our fertility measures and may 
impact mortality. 
To understand these effects, I address two additional questions.

@@

The primary outcomes I examine are average birth intervals, the sex ratio of births, 
and the likelihood of giving birth. 
Because sex selection might also affect the distribution of birth intervals, I furthermore
examine the 25th and 75th percentiles of birth intervals. 
I estimate the model across four periods to capture the changing access and 
legality of sex selection. 
The key variables are maternal education, the sex composition of previous 
children, and area of residence.

@@

% Main results
There are four main results.

First, average birth intervals lengthen over the four decades, more so the higher the 
parity and the more educated the woman.
Women most likely to use sex selection---well-educated with no sons---had the most 
substantial lengthening of birth intervals and the most biased sex ratios. 
As a result, we now see cases that reverse the traditional spacing pattern, with some 
women with no sons having \emph{longer} birth intervals than those with sons.
Those least likely to use sex selection---less-educated in rural areas---still follow 
the traditional pattern of short spacing when they have girls and little evidence of sex 
selection.

Second, the likelihood of a very short birth interval changed little. 
Hence, the increased average spacing comes mainly from longer birth intervals getting even 
longer, primarily because of sex selection. 
The exception is that those better-educated women with a very high predicted use of sex 
selection conceive later and are less likely to have a very short birth interval, precisely 
because their next child is almost certainly a boy.

Third, the fertility rate substantially overestimated how fast cohort fertility fell in the 
1990s and early 2000s as spacing began to increase. 
Although the two have lately been converging, the predicted cohort fertility is still ten to 
twenty percent higher than the fertility rate. 
Furthermore, predicted cohort fertility is still at or above replacement level for all but 
the best-educated urban women.

>>>>

\date{August 2020}

@@

% Demography abstract
\noindent 
Using four rounds of India's National Family and Health Surveys and a
competing-risk hazard model, I show that Hindu women's birth intervals
increased over the last four decades, more so the higher the parity and
the more educated the woman. 
Women most likely to use sex selection---well-educated with no sons---had the most 
substantial lengthening of birth intervals and the most biased sex ratios. 
However, the likelihood of very short spacing did not change substantially. 
As a result of the longer spacing, fertility rates significantly
overestimated how fast cohort fertility fell. 
Although cohort fertility
and the fertility rate have started to converge, the cohort fertility is
still substantially higher than the fertility rate. 
Furthermore, cohort fertility is still at or above replacement level for all but the
best-educated urban women. 
Finally, infant mortality risk has declined substantially over time for all groups, but 
fastest for the lower education groups, who are now close to the level of women with the most
education. 
Short birth spacing is still associated with higher mortality, although substantially less 
so for the best-educated women. 
There is no evidence that the increasing use of sex selection is associated with
higher infant mortality risk.
\noindent JEL: J1, O12, I1
\noindent Keywords: India, prenatal sex determination, censoring, competing risk, non-proportional hazard
\end{abstract}

@@

% Questions addressed and why valuable
The main question here is: how did birth spacing respond to these changes, especially 
the spread of sex selection? 
The motivation is twofold.
First, prior research has failed to appreciate that each abortion increases the
interval between births by six to twelve months.%
\footnote{
The increase consists of three parts. 
First, after an abortion, the uterus needs at least two menstrual cycles to recover, 
or the likelihood of spontaneous abortion increases substantially \citep{zhou00b}. 
Second, the waiting time to conception is one to six months. 
Finally, sex determination tests are reliable only from three months of gestation. 
}
Hence, the growing use of sex-selective abortions may substantially increase birth spacing, 
but we know little about how much. 
Second, more female education, higher household income, and low and declining female labor 
force participation all likely impact birth spacing. 
Hence, the combined changes in birth spacing may outpace what we have observed in 
other countries.

Studying birth spacing contributes to our understanding of fertility decisions, but, equally 
important, birth spacing affects the reliability of our fertility measures and may 
impact mortality. 
I, therefore, address two additional questions.

@@

The primary outcomes I examine are 25th, 50th, and 75th percentile birth intervals, the 
sex ratio at birth, and the likelihood of giving birth. 
I estimate the model across four periods to capture the changing access and 
legality of sex selection. 
The key variables are maternal education, the sex composition of previous 
children, and area of residence.

@@

% Main results
There are three main results.

First, birth intervals lengthen over the four decades, more so the higher the parity, the 
more educated the woman, and the higher the percentile. 
Women most likely to use sex selection---well-educated with no sons---had the most 
substantial lengthening of birth intervals and the most biased sex ratios. 
As a result, we now see cases that reverse the traditional spacing pattern, with some 
women with no sons having \emph{longer} birth intervals than those with sons. 
Those least likely to use sex selection---less-educated in rural areas---still follow the 
traditional pattern of short spacing when they have girls and only some evidence of sex 
selection. 
The likelihood of a very short birth interval changed little.

Second, the fertility rate substantially overestimated how fast cohort fertility fell in the 
1990s and early 2000s as spacing began to increase. 
Although the two have lately been converging, the predicted cohort fertility is still ten to 
twenty percent higher than the fertility rate. 
Furthermore, predicted cohort fertility is still at or above replacement level for all but 
the best-educated urban women.
